\section{Möbius transformations}

In the following we define the group of Möbius transformations and collect some useful basic properties.

\begin{definition}
\label{dfn_MoebiusTransform}
\index{Mobius transformation@Möbius transformation}
A non-constant rational function $\phi \in \C(z)$ of the form 
\begin{equation*}
\phi(z) = \moebius{a}{b}{c}{d}{z},\quad a, b, c, d \in \C,\quad ad - bc \ne 0
\end{equation*}
is called \emph{Möbius transformation}.
\end{definition}

\begin{remark}
The condition $ad - bc \ne 0$ just ensures that $\phi$ is in fact non-constant.
\end{remark}

\begin{theorem}
\label{thm_MoebiusGroup}
The set of Möbius transformations forms a group under the action of function composition and can be identified with the projective general linear group $\PGL{\C}$ or the projective special linear group $\PSL{\C}$.
\end{theorem}
\begin{proof}
Let $\phi$ and $\psi$ be Möbius transformations with
\begin{equation*}
\phi(z) = \moebius{a}{b}{c}{d}{z},\quad \psi(z) = \moebius{e}{f}{g}{h}{z}.
\end{equation*}

First, we make the trivial observation, that composing those two transformations again yields a rational function of the desired form:
\begin{equation}
\label{eqn_MoebiusComposition}
\phi \circ \psi(z) 
 = \moebius{a}{b}{c}{d}{\moebius{e}{f}{g}{h}{z}} 
 = \frac{aez + af + bgz + bh}{cez + cf + dgz + dh} 
 = \moebius{(ae + bg)}{(af + bh)}{(ce + dg)}{(cf + dh)}{z}
\end{equation}

Having a closer look on the resulting coefficients one might notice that they relate to the following matrix product:
\begin{equation}
\label{eqn_MatrixProduct}
\mat{a}{b}{c}{d} \cdot \mat{e}{f}{g}{h} 
 = \mat{ae + bg}{af + bh}{ce + dg}{cf + dh}
\end{equation}

This motivates the definition of a mapping $\pi$ between matrices in $\GL{\C}$ and Möbius transformations:
\begin{equation}
\label{eqn_homPi}
\pi: \mat{a}{b}{c}{d} \mapsto \left(z \mapsto \moebius{a}{b}{c}{d}{z}\right)
\end{equation}
Note, that the domain of $\pi$ is $\GL{\C}$, \ie the set of 2-by-2 matrices with nonzero determinant. This is perfectly consistent with the condition $ad - bc \ne 0$ we have for Möbius transformations. For this reason, $\pi$ is a well-defined function from $\GL{\C}$ to the set of Möbius transformations.

But $\pi$ is not just a function, it is in fact a homomorphism, as we have seen in (\ref{eqn_MoebiusComposition}) and (\ref{eqn_MatrixProduct}). Trivially, $\pi$ is also surjective, which carries over the group structure of $\GL{\C}$ to the set of Möbius transformations. The kernel of $\pi$ comprises of all multiples of the identity matrix. Therefore, by the first isomorphism theorem the group of Möbius transformations is isomorphic to $\GL{\C}/\ker{\pi} \cong \PGL{\C}$ and we have seen in Example \ref{ex_ProjAndGenLinGrp} that $\PGL{\C} \cong \PSL{\C}$.
\end{proof}

\begin{remark}
We note that the nature of Möbius transformations is threefold: Firstly, as in Definition~\ref{dfn_MoebiusTransform}, we can regard a Möbius transformation $\phi$ as purely algebraic object, namely as rational function, \ie the (formal) quotient of two polynomials in $\C[z]$:
\begin{equation*}
\phi_{\text{alg}} = \moebius{a}{b}{c}{d}{z} \in \C(z).
\end{equation*} 
Secondly, $\phi$ has a natural interpretation as meromorphic function on the extended complex plane $\EC = \C \cup \{\infty\}$ in the sense of complex analysis:
\begin{equation*}
\phi_{\text{fun}}: \left\{ 
\begin{array}{ccc}
\EC &\to& \EC \\
z  &\mapsto& \moebius{a}{b}{c}{d}{z}
\end{array}
\right.
\end{equation*}
In a more formal context, this correspondence can also be seen the following way: The group of Möbius transformations acts on the set $\EC$ in the sense of Defintion~\ref{dfn_GroupAction} by $\phi_{\text{alg}} z := \phi_{\text{fun}}(z)$. Now, the homomorphism of Theorem~{\ref{thm_GroupActionHom}} is in fact an isomorphism which assigns each Möbius transformation $\phi_{\text{alg}}$ a permutation of the set $\EC$ which is exactly the meromorphic function $\phi_{\text{fun}}$.
Last but not least, we have seen in Theorem~\ref{thm_MoebiusGroup} that we can also regard $\phi$ as equivalence class of matrices:
\begin{equation*}
\phi_{\text{lin}} = \mat{a}{b}{c}{d}_{\sim} \in \PGL{\C}.
\end{equation*}
Whenever there is no danger of confusion, we will from now on switch between these different views on Möbius transformations seamlessly and exploit concepts of algebra, function theory and linear algebra alternately.
\end{remark}

\begin{lemma}
\label{lem_MoebiusGenerators}
The group of Möbius transformations is generated by the following basic types of transformations:

\begin{tabular}{r l l l}
\index{Translation}
\index{Rotation}
\index{Dilation}
\index{Inversion}
$\bullet$ & Translations: & $z \mapsto z + \alpha$         & $\alpha \in \C$ \\
$\bullet$ & Dilations:    & $z \mapsto \rho z$             & $\rho > 0$ \\
$\bullet$ & Rotations:    & $z \mapsto \epo{\ii \theta} z$ & $\theta \in (-\pi,\pi]$ \\
$\bullet$ & Inversion:   & $z \mapsto \reci{z}$           & ~ 
\end{tabular}
\end{lemma}
\begin{proof}
Let $\phi(z) = \moebius{a}{b}{c}{d}{z}$ be an arbitrary Möbius transformation. In the case when $c = 0$, we may further assume w.l.o.g. that $d = 1$ such that the transformation simply writes $\phi(z) = a z + b$. Obviously this is dilation and rotation by the factor $a$ followed by translation by $b$.

Let's consider the more interesting case, when $c \ne 0$. Without restriction we may assume that $c = 1$, such that 
\begin{equation*}
\phi(z) = \moebius{a}{b}{}{d}{z} = a + \frac{b - ad}{z + d}.
\end{equation*}
Also in this case it is easy to see that $\phi$ is composed of translation by $d$, inversion, dilation and rotation by the factor $b - ad$ and a final translation by $a$.
\end{proof}

Now that we have defined the group of Möbius transformations, it is worth to get a better geometric intuition about how these maps act on the complex plane. Lemma \ref{lem_MoebiusGenerators} gives a first insight, as translations, dilations and rotations are quite easy to understand. Also the map $z \mapsto \reci{z}$ has a geometric interpretation, namely as circle inversion followed by a reflection.

\index{Circle inversion}
\index{Inversion}
In 2-dimensional geometry, \emph{circle inversion} with respect to a reference circle with center $C$ and a radius $r$ takes each point $P$ on the plane to a point $P^{\prime}$ which is determined by $CP \cdot CP^{\prime} = r^2$. The image of $C$ is defined to be the point at infinity (and vice versa). Roughly speaking, the inversion turns the circle ``inside out'', \ie points inside the reference circle are bijectively mapped to points outside while rays from the circle center are invariant under the circle inversion.

\todo{15}{Figure for circle inversion}

Coming back to the concrete map $z \mapsto \reci{z}$, it can now be interpreted the following way: Circle inversion with regard to the unit circle $\UnitCirc$ maps each $z \in \C$ to $\frac{z}{\abs{z}^2} = \reci{\conj{z}}$. Then, reflection across the real axis (\ie complex conjugation) takes $\reci{\conj{z}}$ to $\reci{z}$. 

Summing up, all the basic types of Möbius transformations mentioned in Lemma \ref{lem_MoebiusGenerators} have a very direct geometric interpretation. Still, arbitrary Möbius transformations (especially those involving at least one inversion) are hard to describe in a similar geometric and intuitive way. 

Luckily there is another characterization of Möbius transformations which is both, elegant and visually accessible.

% ---------------------------------------- Subsection: Stereographic projection
\input{sub-intro-mreveal.tex}

% --------------------------------------------- Subsection: Generalized circles
\input{sub-intro-gencirc.tex}

