\section{Groups and basic algebraic constructions}

In this section, we will recapitulate some basic algebraic concepts, most important the notions of \emph{free groups} and \emph{free products}, and the construction of a group in terms of \emph{generators} and \emph{relations}. Additionally, we give the definitions of the \emph{(projective) general} and \emph{special linear groups}. Lastly, \emph{group actions} on sets will be introduced. A reader familiar to this concepts may readily skip this section. Moreover, we will give no rigorous proofs here, as they may be found in many algebra books, as for example in \Hungerford{}.

\subsection{Groups and homomorphisms}

\index{Binary operation}
If $G$ is a nonempty set, a \emph{binary operation} on $G$ is a function $G \times G \to G$. Commonly used notations for the image of $(a,b) \in G \times G$ under a binary operation are $a \cdot b$ or $ab$ (product notation), $a + b$ (additive notation), $a \circ b$, $a \ast b$, etc. In this chapter and also later on we will use the \emph{product notation}, $(a,b) \mapsto a \cdot b =  ab$, most frequently and we refer to $ab$ as the \emph{product} of $a$ and $b$.

\begin{definition}[Groups and monoids]
\label{dfn_GroupMonoid}
\index{Group}
\index{Monoid}
Let $G$ be a nonempty set together with a binary operation on $G$. If the binary operation satisfies the following three axioms,
\index{Associativity}
\begin{enumerate}[(i)]
\item \label{itm_Associativity} \emph{Associativity:} $\forall a,b,c \in G: a(bc) = (ab)c$,
\item \label{itm_IdentityExists} \emph{Existence of an identity element:} $\exists e \in G\ \forall a \in G: ea = ae = a$,
\item \emph{Existence of inverse elements:} $\forall a \in G\ \exists \inv{a} \in G: a \inv{a} = \inv{a} a = e$,
\end{enumerate}
then $G$ is called a \emph{group}. If the axioms (\ref{itm_Associativity}) and (\ref{itm_IdentityExists}) are satisfied, then $G$ is called a \emph{monoid}. Of course every group is in particular a monoid. The notation $\langle G,\cdot,e \rangle$ will be used for making the identity element and the involved binary operation explicit.
\end{definition}

It is convenient to introduce the following notations operating on subsets of a group $G$. For $g \in G$ and $A,B \subseteq G$ we define
\begin{eqnarray*}
\inv{A} &:=& \setdef{\inv{a}}{a \in A} \subseteq G, \\
AB &:=& \setdef{a b}{a \in A, b \in B} \subseteq G,\\
gA &:=& \{g\}A = \setdef{g a}{a \in A} \subseteq G,\\
Ag &:=& A\{g\} = \setdef{a g}{a \in A} \subseteq G.
\end{eqnarray*}

\begin{definition}[Subgroups]
\label{dfn_Subgroup}
\index{Subgroup}
Let $\langle G,\cdot,e \rangle$ be a group. A nonempty subset $H \subseteq G$ which is closed under the binary operation $\cdot$ and inversion, \ie $H\inv{H} = H$, is called a \emph{subgroup} of $G$, in symbols $H \le G$. In particular, $\langle H,\cdot,e \rangle$ is itself a group. If $H \le G$ and $H \ne G$, $H$ is called a \emph{proper subgroup} of $G$, and we write $H \lneq G$. The subgroup $\{e\} \le G$ is called the \emph{trivial subgroup}.
\end{definition}

\begin{definition}[Cosets]
\label{dfn_Coset}
\index{Group!coset}
\index{Coset}
Let $H$ be as subgroup of a group $G$ and $g \in G$. The set $gH$ is called a \emph{left coset} of $H$ in $G$ and $Hg$ is called a \emph{right coset}.
\end{definition}

\index{Conjuagtion}
If $H$ is a subgroup of $G$, then for every $g \in G$ also $gH\inv{g}$ is a subgroup of $G$. For $h \in H$, $g \in G$, we call the operation $h \mapsto g h \inv{g}$ \emph{conjugation} of $h$ by $g$.

\begin{definition}[Normal subgroups]
\label{dfn_NormalSubgroup}
A subgroup $N$ of a group $G$ which is closed under conjugation by elements of $G$, \ie $g N \inv{g} = N$ for all $g \in G$, is called a \emph{normal subgroup} of $G$, in symbols $N \unlhd G$.
\end{definition}

If $N$ is a normal subgroup of $G$, then we have $gN = Ng$ for every $g \in G$, in other words left and right cosets of $N$ in $G$ coincide. We can therefore define on the set of cosets
\begin{equation}
\label{eqn_FactorGroup}
G / N := \setdef{gN}{g \in G},
\end{equation}
a binary operation by $aNbN := abNN = abN$ for $a,b \in G$. It is easy to see that $G / N$ together with this operation forms a group.

\begin{definition}[Factor groups]
\label{dfn_FactorGroup}
Let $N$ be a normal subgroup of a group $G$. The set $G / N$ defined in (\ref{eqn_FactorGroup}) together with the binary operation $(aN,bN) \mapsto abN$ is called the \emph{factor group} (or \emph{quotient group}) of $G$ by $N$.
\end{definition}

\begin{definition}[Homomorphisms]
\label{dfn_Homomorphism}
\index{Homomorphism}
\index{Monomorphism}
\index{Epimorphism}
\index{Isomorphism}
\index{Endomorphism}
\index{Automorphism}
Let $G$ and $H$ be groups (or monoids). A function $f : G \to H$ is a \emph{homomorphism} from $G$ to $H$, if and only if
\begin{equation*}
f(ab) = f(a) f(b) \quad \text{for all } a,b \in G.
\end{equation*}
Depending on $f$ being injective and/or surjective, one distinguishes \emph{monomorphisms} ($f$ is injective), \emph{epimorphisms} ($f$ is surjective) and \emph{isomorphisms} ($f$ is bijective). We call the groups $G$ and $H$ \emph{isomorphic}, in symbols $G \cong H$, if and only if there exists an isomorphism $f : G \to H$. Finally, a homomorphism $f : G \to G$ of $G$ to itself is called \emph{endomorphism}. A bijective endomorphism is called \emph{automorphism}.
\end{definition}

\begin{definition}[Kernel]
\label{dfn_Kernel}
Let $G$ and $H$ be groups and denote the identity element of $H$ by $e_H$. For a homomorphism $f: G \to H$, the set
\begin{equation}
\ker f := \setdef{g \in G}{f(g) = e_H}
\end{equation}
is called the \emph{kernel} of f.
\end{definition}
\begin{lemma}
\label{lem_KernelAndImage}
Let $G$ and $H$ be groups and $f : G \to H$ be a homomorphism. The kernel of $f$ is a normal subgroup of $G$ and the image $f(G)$ is a subgroup of $H$:
\begin{equation*}
\ker f \unlhd G \quad \text{and} \quad f(G) \le H.
\end{equation*}
\end{lemma}
\begin{proof}
Denote the identity elements of $G$ and $H$ by $e_G$ and $e_H$ respectively. Because $f$ is a homomorphism, we have $f(e_G) = f(e_G e_G) = f(e_G) f(e_G)$. Multiplying this by $\inv{f(e_G)}$ yields $e_H = f(e_G)$. In other words $e_G \in \ker f$ and $\ker f$ is nonempty. To see that $\ker f \le G$, we need to show that for all $a, b \in \ker f$ also $a \inv{b} \in \ker f$ -- but this is easy:
\begin{equation*}
f(a \inv{b}) = f(a) f(\inv{b}) = f(a) \inv{f(b)} = e_H \inv{e_H} = e_H.
\end{equation*}
To see that $\ker f$ is a normal subgroup of $G$, let $a \in \ker f$, $g \in G$ and observe
\begin{equation*}
f(g a \inv{g}) = f(g) f(a) f(\inv{g}) = f(g) e_H \inv{f(g)} = e_H.
\end{equation*}
For this reason $g a \inv{g} \in \ker f$ and $\ker f$ is closed under conjugation by elements of $G$. For the second part of the proof, let $f(a), f(b) \in f(G)$ with $a,b \in G$. Clearly $f(a) \inv{f(b)} = f(a) f(\inv{b}) = f(a \inv{b}) \in f(G)$ and therefore $f(G) \le H$.
\end{proof}

\begin{theorem}[First isomorphism theorem]
\label{thm_FirstIsoThm}
Let $G$ and $H$ be groups and $f: G \to H$ be a homomorphism. Then the quotient group $G / \ker f$ is isomorphic to $f(G)$:
\begin{equation*}
G / \ker f \cong f(G).
\end{equation*}
\end{theorem}
\begin{proof}
Let us again denote the identity element of $H$ by $e_H$. Define $N := \ker f$ and a map $\phi : G/N \to f(G)$ by $aN \mapsto f(a)$. Now for arbitrary $a,b \in G$ observe
\begin{equation*}
aN = bN \Leftrightarrow
a \inv{b} \in N \Leftrightarrow
e_H = f(a \inv{b}) = f(a) \inv{f(b)} \Leftrightarrow
f(a) = f(b).
\end{equation*}
This shows two things: First ($\Rightarrow$), $\phi$ is well defined and second ($\Leftarrow$), $\phi$ is injective. Clearly $\phi$ is also surjective and by definition we have
\begin{equation*}
\phi(aN) \phi(bN) = f(a) f(b) = f(ab) = \phi(abN) \quad \text{for all } a,b \in G.
\end{equation*}
Therefore $\phi$ is indeed an isomorphism from $G/N$ to $f(G)$.
\end{proof}

\subsection{Free monoids}
\index{Free!monoid}
Let $\Syms$ be a set of formal symbols, for example $\Syms = \{a, b, c, \dots\}$. For $n \in \N$, $n \ge 0$, we denote by $\Syms^n$ the set of tuples of length $n$ over $\Syms$ and we will call these tuples \emph{words of length $n$ over the alphabet $\Syms$}. For simplicity, we will omit parentheses and commas when notating such a word, for example $(a,b,b,a) =: abba \in \Syms^4$ is a word of length 4. Note the special case for $n = 0$. By definition, $\Syms^0$ contains all words of length $0$. Obviously, there is only one such word, the so-called \emph{empty word}, which we denote by $\epsilon$. A second special case occurs for $\Syms = \emptyset$, where $\Syms^n = \emptyset$ for all $n > 0$ and $\Syms^0 = \{\epsilon\}$, because there are no words over the empty alphabet except the empty word. Next, we define 
\begin{equation}
\label{eqn_SigmaStarDef}
\Words{\Syms} := \bigcup_{n \ge 0} \Syms^n.
\end{equation}
as the set of all \emph{words over the alphabet $\Syms$}. It is now just natural to define a binary operation $\cdot$ on $\Words{\Syms}$ which is given by concatenation of words:
\begin{equation*}
w_1 \cdot w_2 := w_1 w_2, \quad w_1, w_2 \in \Words{\Syms}.
\end{equation*}
It is obvious that this operation on $\Words{\Syms}$ is associative and has the empty word $\epsilon$ as identity element. Therefore, according to Definition~\ref{dfn_GroupMonoid}, the algebraic structure $\langle \Words{\Syms}, \cdot, \epsilon \rangle$ forms a monoid.

\begin{definition}[Free monoid]
\label{dfn_FreeMonoid}
Let $\Syms$ be an arbitrary set of symbols (called the alphabet), and $\Words{\Syms}$ be defined as in (\ref{eqn_SigmaStarDef}). Moreover, denote concatenation of words by $\cdot$ and let $\epsilon$ be the empty word. Then the algebraic structure $\langle \Words{\Syms}, \cdot, \epsilon \rangle$ is called the \emph{free monoid over the alphabet $\Syms$}.
\end{definition} 

\subsection{Free groups}
\index{Free!group}
\index{Formal inverse}
In a similar fashion, we can construct also a group from any given formal alphabet $\Syms$. For this purpose we first choose a ``disjoint copy'' of $\Syms$ which we denote by $\inv{\Syms}$. To be precise, $\inv{\Syms}$ may be any arbitrary set satisfying $\Syms \cap \inv{\Syms} = \emptyset$ and having same cardinality as $\Syms$. In particular we could for example set $\inv{\Syms} := \Syms \times \{1\}$. Next, we choose a bijection $f : \Syms \to \inv{\Syms}$. We can extend this bijection between the sets $\Syms$ and $\inv{\Syms}$ to an involution $\overline{f}$ living on the union $\overline{\Syms} := \Syms \cup \inv{\Syms}$ by defining $\overline{f} := f \cup \inv{f}$ or more verbose:
\begin{equation*}
\overline{f}(a) := \begin{cases}
f(a)       & \text{if } a \in \Syms\\
\inv{f}(a) & \text{if } a \in \inv{\Syms}
\end{cases}
\end{equation*}
Now we introduce the notation $\overline{f}(a) =: \inv{a}$ for all $a \in \overline{\Syms}$ and call $\inv{a}$ the \emph{(formal) inverse} of the element $a$. 

\index{Reduced form}
\index{Group!word}
Having now defined a formal inverse for each of our symbols in the given alphabet, we can go on and consider the free monoid $\Words{\overline{\Syms}}$. If $\sigma_1, \sigma_2, \dots, \sigma_n$ are symbols of $\overline{\Syms}$, then we say the word $\sigma_1 \sigma_2 \dots \sigma_n$ is \emph{reduced}, if and only if no two subsequent symbols of the word are inverse to each other, that is
\begin{equation}
\label{eqn_GrpWordReducedForm}
 \sigma_j \ne \sigma_{j+1}^{-1} \quad \text{for all } 1 \le j < n.
\end{equation}
Clearly every word $w \in \Words{\overline{\Syms}}$ can be brought into reduced form by successively ``canceling out'' adjacent inverse symbols until finally a word is obtained, which satisfies (\ref{eqn_GrpWordReducedForm}). We call the result of this procedure the \emph{reduced form of $w$}. Additionally we define two words $w_1, w_2 \in \Words{\overline{\Syms}}$ to be \emph{equivalent} if and only if they have the same reduced form and we write $w_1 \sim w_2$ in this case. If we denote the reduced form of $w$ by $\phi(w)$, we can write
\begin{equation*}
w_1 \sim w_2 \Leftrightarrow \phi(w_1) = \phi(w_2).
\end{equation*}
Moreover we observe that if $w_1 \sim w_2$ and $v_1 \sim v_2$, then we also have $w_1 v_1 \sim w_2 v_2$ because $\phi(w_1) = \phi(w_2) =: w$ and $\phi(v_1) = \phi(v_2) =: v$ implies 
\begin{equation*}
\phi(w_1 v_1) = \phi(w v) = \phi(w_2 v_2).
\end{equation*}
Thus we see that the equivalence relation $\sim$ is compatible with the operation of word concatenation. Therefore we can consider also the set of equivalence classes $\Words{\overline{\Syms}}/{\sim}$ as a monoid under the operation of word concatenation. Obviously the set of reduced words is a system of representatives for $\Words{\overline{\Syms}}/{\sim}$ and we agree to denote an equivalence class $[w]_{\sim}$ simply by the reduced word $w$. $\Words{\overline{\Syms}}/{\sim}$ is not just a monoid, but in fact a group, as the inverse of a word $\sigma_1 \sigma_2 \dots \sigma_{n-1} \sigma_n$ is trivially given by $\sigma_n^{-1} \sigma_{n-1}^{-1} \dots \sigma_2^{-1} \sigma_1^{-1}$. We call a reduced word of $\Words{\overline{\Syms}}$ (resp. its corresponding equivalence class in $\Words{\overline{\Syms}}/{\sim}$) a \emph{group word over the alphabet $\Syms$}.

\begin{definition}[Free group]
\label{dfn_FreeGroup}
Let $\Syms$ be an arbitrary set of formal symbols and define $\overline{\Syms} := \Syms \cup \inv{\Syms}$ as above. On the free monoid $\Words{\overline{\Syms}}$ define an equivalence relation $\sim$ by identifying words with the same reduced form. Then the algebraic structure 
\begin{equation}
\label{eqn_FreeGrp}
\FreeGrp{\Syms} := \langle \Words{\overline{\Syms}}/{\sim}, \cdot, \epsilon \rangle
\end{equation}
is called the \emph{free group over the alphabet $\Syms$}.
\end{definition}

\begin{remark}
In the exceptional case when $\Syms = \emptyset$,  we obtain the trivial group by this construction: If $\Syms = \emptyset$ then we can also choose $\inv{\Syms} = \emptyset$ ($\Syms$ and $\inv{\Syms}$ are then disjoint as required). It follows that also $\overline{\Syms}$ is empty and we end up with the trivial monoid $\Words{\overline{\Syms}} = \{\epsilon\}$, which corresponds to the trivial group.
\end{remark}

\begin{example}
Let $\Syms$ be the singleton set $\{a\}$. Now the free group over $\Syms$ consists precisely of the following group words:
\begin{equation*}
\FreeGrp{\Syms} = \{\epsilon, a, aa, aaa, \dots, \inv{a}, \inv{a}\inv{a},\dots\}.
\end{equation*}
If we now introduce the notation
\begin{equation*}
\epsilon =: a^0, \quad 
\underbrace{a a \dots a}_{k \text{ times}} =: a^k, \quad
\underbrace{\inv{a} \inv{a} \dots \inv{a}}_{k \text{ times}} =: a^{-k},
\end{equation*}
we have $a^n \cdot a^m = a^{n+m}$ for all $n,m \in \Z$ and it is thus evident that the free group over any one-element alphabet is isomorphic to the group $\langle \Z, +, 0 \rangle$. 

Note that this is the only case (besides the trivial one, when $\Syms = \emptyset$), where the free group is commutative. In fact, the free group construction always yields a group with ``richest possible'' structure in the following sense:

\begin{theorem}
\label{thm_FreeGrpUniqueHom}
Let $\Syms$ be an arbitrary set and $G$ be an arbitrary group. If any mapping $\phi : \Syms \to G$ is given, then we can always extend $\phi$ in a unique way to a homomorphism $\overline{\phi}: \FreeGrp{\Syms} \to G$.
\end{theorem}

In fact, the property described in Theorem \ref{thm_FreeGrpUniqueHom} completely characterizes the free group up to isomorphism, as stated in the next Theorem:

\begin{theorem}[Universal mapping property]
\label{thm_FreeGrpUnivMapProp}
Let $\Syms$ be an arbitrary set and $F$ be a group together with an injective map $\iota : \Syms \to F$. If $F$ has the property that for any map $\phi : \Syms \to G$, where $G$ is an arbitrary group, there is a homomorphism $\overline{\phi} : F \to G$ such that $\phi = \overline{\phi} \circ \iota$,
then $F$ is (isomorphic to) the free group over the alphabet $\Syms$.
\end{theorem}

\end{example}

\subsection{Generators and relations}

\begin{definition}
\index{Generator}
\index{Group!generator}
Let $\langle G, \cdot, 1 \rangle$ be a group and $\Syms \subseteq G$ a subset. We say $G$ is \emph{generated by} the elements of $\Syms$, if every element $g \in G$ can be written as a product of elements from $\Syms$ in the following sense:
\begin{equation}
\label{eqn_GrpGenerators}
\forall g \in G\ \exists n \in \N,\ (\sigma_j) \in \Syms^n,\ (k_j) \in \Z^n:\quad 
g = \sigma_1^{k_1} \sigma_2^{k_2} \dots \sigma_n^{k_n}.
\end{equation}
Moreover in this case we call the elements of $\Syms$ \emph{generators} of $G$. 
\end{definition}

\index{Group!relation}
Let us first consider the case when some $g \in G$ can be generated in two  different ways, for example
\begin{equation*}
g = \sigma_1^{k_1} \sigma_2^{k_2} \dots \sigma_n^{k_n} 
  = \tau_1^{\ell_1} \tau_2^{\ell_2} \dots \tau_m^{\ell_m}
\end{equation*}
with $\sigma_j, \tau_j \in \Syms$ and $k_j, \ell_j \in \Z$. This immediately gives
\begin{equation}
\label{eqn_GrpRelation}
\sigma_1^{k_1} \sigma_2^{k_e} \dots \sigma_n^{k_n} \cdot
\tau_m^{-\ell_m} \tau_{m-1}^{-\ell_{m-1}} \dots \tau_1^{-\ell_1} = 1.
\end{equation}
We call the group word occurring on the left hand side of (\ref{eqn_GrpRelation}) a \emph{relation} on $G$. The set of all such possible relations on $G$ forms a normal subgroup of $\FreeGrp{\Syms}$. To see this, let $\phi : \Syms \to G$ be the canonical embedding. According to Theorem~\ref{thm_FreeGrpUniqueHom} there exists a unique extension of $\phi$ to a homomorphism $\overline{\phi}: \FreeGrp{\Syms} \to G$. In fact this homomorphism just evaluates group words over $\Syms$ in the obvious way to concrete group elements in $G$. By definition, the set of all relations on $G$ is now given by $N := \ker(\overline{\phi})$ which is, by Lemma~\ref{lem_KernelAndImage}, a normal subgroup of $\FreeGrp{\Syms}$. It now follows from the first isomorphism theorem (Theorem~\ref{thm_FirstIsoThm}) that $G \cong \FreeGrp{\Syms} / N$. In the case when the product representation in (\ref{eqn_GrpGenerators}) is unique for all $g \in G$, then $N$ just consists of the identity element and $G$ is isomorphic to $\FreeGrp{\Syms}$. In this case $G$ is said to be \emph{relation-free}, which is where the terminology ``free group'' actually comes from. 

Summing up, we see that every group can be described by supplying a set of generators $\Syms$ and a the set of relations $N \unlhd \FreeGrp{\Syms}$. Of course it is also sufficient to supply just a subset $R$ of relations from which the other relations can be derived. This leads to the following definition:

\begin{definition}[Presentation of a group]
\label{dfn_GrpConstructGenRel}
Let $\Syms$ be an arbitrary set and $R \subseteq \FreeGrp{\Syms}$ a set of group words over the alphabet $\Syms$. A group $G$ is said to be the \emph{group defined by} the \emph{generators} $\sigma \in \Syms$ \emph{and relations} $R$, if $G$ is obtained by the following construction:
\begin{enumerate}
\item Let $N$ be the normal subgroup of $\FreeGrp{\Syms}$ generated by the relations $R$:
\begin{equation*}
N := \bigcap_{R \subseteq N^{\prime} \unlhd \FreeGrp{\Syms}} N^{\prime}.
\end{equation*}
\item Define $G := \presentation{\Syms}{R} := \FreeGrp{\Syms}/N$.
\end{enumerate}
Finally $\presentation{\Syms}{R}$ is called a \emph{presentation} of $G$.
\end{definition}

For easier notation we may omit braces when listing $\Syms$ and $R$ in a presentation of $G$. Moreover we may write out relations more explicitly, for example instead of $G = \presentation{\{a\}}{\{a^2\}}$ we may write $G = \presentation{a}{a^2 = 1}$.

\begin{remark}
\index{Word problem}
The presentation $\presentation{\Syms}{R}$ of a group $G$ is by far not unique as the normal group $N$ can be generated by many different subsets $R \subseteq N$. Even worse, the $\emph{word problem}$, \ie the question, if a given group word $w \in \FreeGrp{\Syms}$ is in $N$ when $R$ is given, is in general not decidable, which means that there is no general algorithm for deciding, if two given group words represent the same element of $G$.
\end{remark}

\begin{example}
The group $\presentation{t,r}{t^2 = r^3 = 1}$ consists of all group words of the form
\begin{equation*}
\label{ex_RTGroup}
r^{k_1} t r^{k_2} t \dots t r^{k_n}, \quad 
\text {with } k_1, k_n \in \{0,\pm 1\} \text{ and } k_2, \dots, k_{n-1} \in \{\pm 1\}.
\end{equation*}
\end{example}
We conclude this section with the statement, that among all groups, which satisfy a given set of relations $R$, the group constructed as above is in a certain sense the largest possible one.

\begin{theorem}
Let $\presentation{\Syms}{R}$ be a presentation of a group $G$. If $H$ is any group generated by $\Syms$ and if $H$ satisfies all relations $R$, then there is an epimorphism $G \to H$.
\end{theorem}

\subsection{Free products}

\index{Free!product}
Let $\langle G_i, \cdot, e_i \rangle_{i \in I}$ a family of pairwise disjoint groups, \ie $G_i \cap G_j = \emptyset$ for $i \ne j$. We define the \emph{free product} of the family of groups $(G_i)_{i \in I}$ by a construction in terms of generators and relations (see Definition~\ref{dfn_GrpConstructGenRel}) as follows: As the set of generators we choose the alphabet $\Syms = \bigcup_{i \in I} G_i$ and as the set of relations we take all the relations coming from any of the groups $G_i$. Note that a priori the neutral elements $e_i \in G_i$ are all different symbols, but the final factorization by the normal subgroup $N$ automatically identifies them with each other, as $N$ includes all the $e_i$ as trivial relations.

\begin{definition}[Free product]
Let $(G_i)_{i \in I}$ be a family of disjoint groups. For every $i \in I$, let $\overline{\phi}_i$ be the unique extension of the identity map on $G_i$ to a homomorphism $\FreeGrp{G_i} \to G_i$ according to Theorem~\ref{thm_FreeGrpUniqueHom}.\footnote{In other words, $\overline{\phi}_i$ is the map which evaluates group words over the alphabet $G_i$ to concrete elements of $G_i$ in the obvious way.} The \emph{free product} of the family $(G_i)_{i \in I}$ is defined as
\begin{equation*}
{\prod_{i \in I}}^{\star} G_i := \presentation{\Syms}{R}, 
\quad \text{where } \Syms = \bigcup_{i \in I} G_i \text{ and } R = \bigcup_{i \in I} \ker \overline{\phi}_i.
\end{equation*} 
\end{definition}

If only a small finite number of groups is involved, for example only the two groups $G$ and $H$, we will write $G \ast H$ for their free product.

\begin{example}
\label{ex_RTFreeProd}
The free product of the groups $G = \presentation{t}{t^2 = 1}$ and $H = \presentation{r}{r^3 = 1}$ is the group $G \ast H = \presentation{t,r}{t^2 = r^3 = 1}$ which we have already seen in Example \ref{ex_RTGroup}.
\end{example}

\begin{example}
Let $F_n$ and $F_m$ be the free groups generated by $n$ and $m$ elements respectively. Then the free product $F_n \ast F_m = F_{n+m}$ is the free group generated by $n+m$ elements.
\end{example}

\subsection{Basic linear groups}
In this section we give the definitions of some important basic groups which will be needed throughout this document. We start with the general and special linear group.

\begin{definition}[General linear group]
\label{dfn_GenLinGrp}
\index{General linear group}
\index{GLnF@$\GLn{n}{F}$}
Let $F$ be a field and $n > 0$. The group of invertible $n$-by-$n$ matrices over  $F$ is called \emph{general linear group} and is denoted by
\begin{equation}
\label{eqn_GeneralLinearGroup}
\GLn{n}{F} := \setdef{M \in \SqMat{F}{n}}{\det M \ne 0}.
\end{equation}
\end{definition}

\begin{definition}[Special linear group]
\label{dfn_SpLinGrp}
\index{Special linear group}
\index{SLnR@$\SLn{n}{R}$}
Let $R$ be a commutative ring with 1 and $n > 0$. The group of $n$-by-$n$ matrices over $R$ having determinant 1, is called \emph{special linear group} and is denoted by
\begin{equation}
\label{eqn_SpecialLinearGroup}
\SLn{n}{R} := \setdef{M \in \SqMat{R}{n}}{\det M = 1}.
\end{equation}
Of course in the case, when $R$ is contained in the field $F$, then the special linear group $\SLn{n}{R}$ is a subgroup of the general linear group $\GLn{n}{F}$. 
\end{definition}

\index{Group!center}
For both, the general and special linear group, one can construct the \emph{projective groups} by identifying matrices which differ by a scalar multiple. For this purpose we define the \emph{center} of a group $G$ in the usual way:
\begin{equation}
\Center{G} := \setdef{z \in G}{zg = gz \quad \forall g \in G}.
\end{equation}
If $G$ is a matrix group, $\Center{G}$ consists precisely of all scalar multiples of the identity matrix within $G$, which is exactly what we need for this construction.

\begin{definition}[Projective linear groups]
\label{dfn_ProjLinGrp}
\index{Projective!general linear group}
\index{PGLnF@$\PGLn{n}{F}$}
\index{Projective!special linear group}
\index{PSLnR@$\PSLn{n}{R}$}
As above, let $F$ be a field, $R$ a ring with 1 and $n > 0$. The \emph{projective general linear group} and the \emph{projective special linear group} are defined as
\begin{IEEEeqnarray}{rClCl}
\label{eqn_ProjGenLinGrp}
\PGLn{n}{F} &:=& \GLn{n}{F} &/& \Center{\GLn{n}{F}}, \IEEEyesnumber \\
\label{eqn_ProjSpLinGrp}
\PSLn{n}{R} &:=& \SLn{n}{R} &/& \Center{\SLn{n}{R}}. \IEEEyesnumber
\end{IEEEeqnarray}
\end{definition}

\begin{example}
\label{ex_ProjAndGenLinGrp}
Consider a matrix $M \in \SL{\R}$. The equivalence class $[M]_{\sim} \in \PSL{\R}$ consists precisely of the two matrices $M$ and $-M$, whereas the equivalence class $[M]_{\sim} \in \PGL{\R}$ is just the set $\setdef{\lambda M}{\lambda \in \R \setminus \{0\}}$. For a clearer distinction of these equivalence classes let us introduce the notation 
\begin{eqnarray*}
{}_{\pm}M     &:=& [M]_{\sim} \in \PSL{\R} \quad \text{and} \\
{}_{\lambda}M &:=& [M]_{\sim} \in \PGL{\R}.
\end{eqnarray*}
Clearly ${}_{\lambda}M = {}_{\lambda}(-M)$, which allows us to define an injective homomorphism $\iota$ which embeds $\PSL{\R}$ into $\PGL{\R}$:
\begin{equation*}
\iota : \left\{\begin{array}{rcl}
\PSL{\R}  &\to&     \PGL{\R} \\
{}_{\pm}M &\mapsto& {}_{\lambda}M
\end{array}\right.
\end{equation*}
Because $\det(\lambda M) = \lambda^2 \det(M) = \lambda^2 > 0$, the image of $\iota$ consists of equivalence classes of matrices with positive determinant only, whereas $\PGL{\R}$ also contains equivalence classes of matrices with negative determinant. Therefore the embedding $\iota$ is not surjective and $\PSL{\R}$ isomorphic to a proper subgroup of $\PGL{\R}$:
\begin{equation*}
\PSL{\R} \cong \iota(\PSL{\R}) \lneq \PGL{\R}.
\end{equation*}
In contrast to that, if we switch to $\PSL{\C}$ and $\PGL{\C}$ respectively and define $\iota$ analogously, we see that for every equivalence class $[M]_{\sim} \in \PGL{\C}$, the matrix $M^{\prime} := \reci{\sqrt{\det M}} M$ is in $\SL{\C}$ and satisfies $\iota([M^{\prime}]_{\sim}) = [M]_{\sim}$. Therefore $\iota$ is now also surjective and thus we have 
\begin{equation*}
\PSL{\C} \cong \PGL{\C}.
\end{equation*}
More generally, for an arbitrary field $F$ the groups $\PSLn{n}{F}$ and $\PGLn{n}{F}$ are isomorphic, if and only if $F$ is closed under taking the $n$-th root, otherwise $\PSLn{n}{F}$ is isomorphic to a proper subgroup of $\PGLn{n}{F}$.
\end{example}

\subsection{The action of a group on a set}

\begin{definition}[Group action]
\label{dfn_GroupAction}
\index{Group!action}
Let $\langle G, \cdot, 1 \rangle$ be a group and $\mathcal{S}$ be an arbitrary set. A function $G \times \mathcal{S} \to \mathcal{S}$, $(g, x) \mapsto g x$ is called a \emph{group action} of $G$ on the set $\mathcal{S}$ (alternatively we say $G$ \emph{acts on the set} $\mathcal{S}$), if and only if
\begin{equation}
\label{eqn_GroupAction}
\forall x \in \mathcal{S},\ g_1, g_2 \in G:\quad 1 x = x \quad \text{and} \quad (g_1 g_2) x = g_1 (g_2 x).
\end{equation}
\end{definition}

\begin{theorem}
Let $G$ be a group acting on the set $\mathcal{S}$.
\begin{enumerate}[(i)]
\item The relation $\sim$ on $\mathcal{S}$ defined by
\begin{equation}
\label{eqn_GroupActionEquivalence}
x_1 \sim x_2 \quad:\Leftrightarrow\quad \exists g \in G: x_1 = g x_2
\end{equation}
is an equivalence relation.
\item For all $x \in \mathcal{S}$ the set $G_x := \setdef{g \in G}{gx = x}$ forms a subgroup of $G$.
\end{enumerate}
\end{theorem}

\begin{definition}[Orbit, stabilizer]
\label{dfn_OrbitStabilizer}
\index{Group!orbit}
\index{Orbit}
\index{Group!stabilizer}
\index{Stabilizer}
For $x \in \mathcal{S}$, the set 
\begin{equation*}
Gx := \setdef{gx \in \mathcal{S}}{g \in G},
\end{equation*}which is identical to the equivalence class $[x]_\sim \in \mathcal{S}/\sim$ of the relation $\sim$ defined in (\ref{eqn_GroupActionEquivalence}), is called the \emph{orbit} of $x$ under $G$. The group 
\begin{equation*}
\Stabilizer{G}{x} := \setdef{g \in G}{gx = x} \le G
\end{equation*} is called the \emph{stabilizer} of $x$.
\end{definition}

\begin{theorem}
\label{thm_GroupActionHom}
The action of a group $G$ on a set $\mathcal{S}$ induces a homomorphism from $G$ to the group of permutations on $\mathcal{S}$ which is given by $g \mapsto (x \mapsto gx)$.
\end{theorem}

\begin{corollary}[Cayley]
The action of a group $G$ on itself induces a monomorphism from $G$ to the group of permutations of its elements. Hence every group is isomorphic to a permutation group. In particular every finite group $G$ is isomorphic to a subgroup of the symmetric group $S_n$ with $n = |G|$.
\end{corollary}
